Microplatforms: Product platforms as code

### 2018-05-31 paper.tex
Small edits for better flow

When building and delivering a real-world client-server application or service, it is important to consider all the parts that form the product, and without which the product would not operate. Typically we think of the product as two major components - the application itself and the platform it runs on. However, there are many other tools, processes, activities and concerns involved in delivering a Software as a Service (SaaS) product to its users than is covered by the functional features and the runtime environment alone (See figure \ref{fig:concerns}).

@@

Continuous Deployment is a method in which every change to the codebase gets deployed immediately, which necessitates a fully automated process of testing and deployment, in order to reduce the risk of human error and speed the process up enough so that it doesn’t dominate all the development activities. The result has the benefit of deploying smaller increments, which have far smaller impact when things go wrong, and significantly reduces the scope of diagnosing the failure and the time to recovery. Delivering product in smaller increments also reduces time to market with the first increment already providing value and customer feedback (and potentially generating revenue), while the rest is being developed.

Typically, when following the continuous deployment method, businesses do not want customers to immediately see the change caused by every deployment and so mechanisms such as feature flags~\cite{FeatureT95:online} are introduced in order to decouple deployment of code changes from releases of customer functionality. This has numerous benefits for both sides, since the ability to release is no longer dependent on the ability to deploy and vice versa.

A platform supporting a digital product which is available to customers any time of day, seven days a week also has to have the ability to tolerate failure of infrastructure or individual components without failure of the product itself. This is commonly referred to as high-availability~\cite{HighAvai70:online} and is generally achieved by running several instances of the same services spread across geographical locations, so that the application can keep working even when an entire data center is lost. This horizontal replication is also used for scaling the application with customer demand~\cite{WhatIsAm85:online}.

An important consideration for a product platform is the cost of building it and the ongoing maintenance involved, including infrastructure provisioning, maintenance (security patches) and scaling capacity to support the products that need to run on it. The effort involved should not be under-estimated~\cite{saaspdf24:online} and is the motivation behind most digital product now running on cloud infrastructure. We will discuss the reasons for this cost reduction in more detail in sections~\ref{sec:cost} and~\ref{sec:automation}.

There has been a clear trend over the past two decades of automating as much of the above activities as possible and treating them as another part of the product itself. This is the core proposition of cloud services and a key concept of the DevOps movement~\cite{DevOpsCu34:online}.

In this paper we will investigate ways of leveraging automation to design, build and manage everything involved in building and deploying digital product to the level where the entire product platform can be fully owned by a small, cross-functional product delivery team, enabling significantly increased capability and velocity of product delivery.

In section \ref{sec:benefits} we outline the goals a good product platform should aim to support. In section \ref{sec:cost} the current approach of using a shared platform is discussed and an alternative (Microplatforms) is suggested in section \ref{sec:microplatforms}, supported by the concept of managing all aspects of it as source code in section \ref{sec:ascode}. Finally, we describe a reference architecture of an example Microplatform in Appendix~\ref{sec:architecture}.

@@

A product platform should ideally support several goals of the organisation in order to realise return on the investment of designing and building it in the first place. The main benefit of a product platform should be reducing the time spent on manual operational tasks, deployments and testing, ultimately translating to increased reliability and greater velocity of the teams. A platform should support the cross-functional team in practicing DevOps: building the product features and delivering them to customers quickly and with minimum risk of causing defects or outages.

The goal of Agile software delivery methodologies like Lean~\cite{LeanSoft20:online} is to eliminate all waste from the process, thus increasing throughput. Focusing on the platform, the waste can be in, for example, being impeded by platform downtime, working around (missing) platform features, repeatedly performing manual tasks or coordinating with other teams to perform tasks (we will discuss the coordination costs in the next section in more detail).

@@

@@

Sharing the platform implicitly requires multi-tenancy features, e.g. the concept of organisations and projects, granular role-based access control, resource limits and isolation, etc. Moreover, in order for the platform to be shared by all teams, the feature set needs to be a union of all the features required by each team. In the theoretical extreme cases, this means supporting $n$ disjoint sets of features for $n$ teams, or forcing all the teams to use the single set of features prescribed by the platform.

@@

After the initial build is completed, more investment needs to be made on an ongoing basis to run operations, i.e. provision new hardware, patch operating systems with fixes, add more resources, clean up older instances of services, etc. In order to serve the business for several years, new features need to be added as they become available in the industry in order to stay within the recognised best practice (often referred to as evergreening).

By far the largest ongoing cost to the business, which is often not explicitly recognised, is the cost of coordination between the product teams and the platform engineering team. Unless every common task from provisioning and getting access to regular deployments and elastic scaling is automated, the product teams need to ask for help from the platform team. The platform team quickly forms a large backlog of day to day tasks to perform, product teams end up queueing behind others, and the platform team have much less time to invest into improvements of the platform itself and automation of the daily tasks.

@@

The overall cost of building and operating a shared product platform is large enough to pose a serious risk to the business. Typically, when things do not go well, this leads to resistance to cut the losses - the sunk cost fallacy~\cite{TheSunkC23:online}. Instead of trying again, or exploring different solutions, the attitude turns into “we will make it work” and results in products being built on a mediocre platform with reliability and scaling issues, increasing risk of preventable outages and slowing down product delivery. This is the final, often unrecognised, cost of a shared product platform - the missed opportunity cost of not delivering as much customer value as possible.

@@

The common theme and, in our opinion, the underlying source of the shared platform costs is the fact they are \textit{shared} among a relatively large number of different product teams. (Notice that the problems listed above seem much less significant if you imagine the platform is only supporting three product teams). The shared nature is what leads to the need for the extra features that enable the sharing, it is also the source of scaling and reliability challenges and, most significantly leads to the rise of coordination costs.

@@

This is not a new finding, automating certain tasks is an accepted practice in multiple industries including software engineering. An obvious example is regression testing using automated unit tests, integration tests and end-to-end tests. Testing used to be a manual task performed by humans following test scripts. Automating it significantly decreased the time between deployments allowing us to get from deploying every few months to several times a day, enabling a completely new working practice known as Continuous Deployment to Production (CDP). This decreases risk of deployments to a minimum due to the size of deployments and reliability of the process. It also minimises mean time to recovery from incidents (requiring a code change to resolve), and results in an overall increase in reliability and a leap in both delivery velocity and ability to adapt to customer needs.

Similarly, the success of cloud services is in a huge part down to allowing automation. Instead of spending weeks manually provisioning servers and network infrastructure, the same infrastructure can be provisioned automatically within minutes. Leveraging cloud services, a small team of engineers can set up a web or mobile application infrastructure that will vastly outperform one set up manually (over weeks or months) by a traditional IT organisation in every measure. [ref?]

@@

An important consequence of the way modern cluster management software is configured is the ability to provide \textbf{service orchestration as code}, i.e. the state of the cluster is configured down to the desired microservices that should run on it and their names, number of replicas and deployment strategies. Coupled with service discovery, this makes it significantly easier to operate applications built with a microservices architecture.

@@

@@

It removes the coordination cost of all product teams depending on a single platform team. If a team needs to change some aspect of their platform or add a missing feature, they can do so for themselves. A platform team will most likely still exist, but the product it delivers is the design and automation of a platform, not the infrastructure itself. This team can become an owner and steward of the platform design, which can and should be internally open source, so that everyone can contribute and the platform engineers can transition into expert support and innovation roles, rather than day to day operations. This is a tried and tested approach which works well for shared, reusable code libraries and services []. With the shared dependency removed, the organisation becomes more horizontally scalable, just like software systems do. Vertical integration leads to horizontal scalability.

@@

Targeting a cluster orchestration technology, applications become fully portable between any infrastructure supporting the technology from a developer’s laptop through any public cloud provider to on-premise solutions. Kubernetes [] is rapidly becoming a de-facto standard in cluster orchestration: it is possible to run a local cluster using a tool like Minikube [], clusters can be provisioned using simple compute resources in different clouds with Kops [], all three major cloud providers are offering or have announced hosted services (GKE [], EKS [] and AKS []) and managed on-premise solutions also exist, e.g. OpenShift by RedHat []or Pivotal Kubernetes Service [].

Finally, the level of automation required for a small platform to be managed by an average sized product team enables a capability crucial to agile software delivery: continuous deployment to production. With continuous deployment to production, every change to the application and infrastructure goes through thorough automated testing and validation. Once the validation passes without issues, the change is deployed to production. This can happen several times a day, even multiple times an hour.

@@

Immutability, i.e. preventing things from being updated in place implicitly creates a version history of all changes and makes it impossible for anything to change unexpectedly or get into a certain state in any other way than being recreated. Removing previous versions of things is a case of garbage collection to reclaim resources they consume. Immutability decouples values, i.e. the entities (infrastructure, build artifacts, deployments, etc.) from the identities, the way we refer to them over time, e.g. “the customer service”. [Rich Hickey]

@@

With immutability and with content addressing especially, it is much easier and efficient to compare two items of the same type (i.e. just by their names), enabling idempotence: the concept of actions which can be repeated many times producing no further result beyond the first application. Enabling easy idempotence leads to huge simplification, as we can typically stop focusing on detail and let idempotence handle the unwanted repetition. In a sense, idempotence unifies creation and update and makes both operations the same from outside, moving solutions from imperative to declarative. We can say that idempotence decouples intent from current state. This also appears multiple times: infrastructure provisioning with Terraform, Docker image builds and pushes to registry, deployment to a cluster or response to failure with the reconciliation loop. Everything tries to reach a declared state.

>>>>

When building and delivering a real-world client-server application or service, it is important to consider all the parts that form the digital product, and without which the product would not operate. Typically we think of the product as two major components - the application itself and the \"platform\" it is hosted on. However, there are many other tools, processes, activities and concerns involved in delivering a Software as a Service (SaaS) product to its users than is covered by the functional features and the runtime environment alone (See figure \ref{fig:concerns}).

@@

A digital product which is available to customers any time of day, seven days a week also has to be built with the ability to tolerate failure of infrastructure or individual components without failure of the product itself. This is commonly referred to as high-availability~\cite{HighAvai70:online} and is generally achieved by running several instances of the same services spread across geographical locations, so that the application can keep working even when an entire data center is lost. This horizontal replication is also used for scaling the application with customer demand~\cite{WhatIsAm85:online}.

Given the scope and complexity of all the activities we just described, it makes economic sense to design a set of processes and tools used to perform and support them and bundle them into a reusable, commodity solution we will refer to as a product platform. The effort involved should not be under-estimated~\cite{saaspdf24:online} and is the chief reason why most large organisations choose the approach of hosting their digital product portfolio on a single centrally managed shared platform. Most organisations also opt to run the product platform on cloud infrastructure, which has prooven to be more flexible and significantly cheaper to operate than the alternative of using on-premise infrastructure. (We will discuss the reasons for this cost reduction in more detail in sections~\ref{sec:cost} and~\ref{sec:automation}.)

There has been a clear trend over the past two decades of automating as much of the above activities as possible and treating them as another part of the product itself, managing them within the same teams as the product engineering. This is the core proposition of cloud services and a key concept of the DevOps movement~\cite{DevOpsCu34:online}. In this paper we will investigate ways of leveraging automation to design, build and manage everything involved in building and deploying digital product to the level where the entire product platform can be fully owned by a small, cross-functional product delivery team, enabling significantly increased capability and velocity of product delivery.

In section \ref{sec:benefits} we outline the goals a good product platform should aim to support. In section \ref{sec:cost} the current approach of using a shared platform is discussed and an alternative (Microplatforms) is suggested in section \ref{sec:microplatforms}, supported by the concept of managing all aspects of it as source code in section \ref{sec:ascode}. Finally, we describe a reference architecture of an example Microplatforms system in Appendix~\ref{sec:architecture}.

@@

A product platform should ideally support as many goals of the organisation as possible in order to realise return on the investment of designing and building it in the first place. The main benefit of a product platform should be reducing the time spent on manual operational tasks, deployments and testing, ultimately translating to increased reliability and greater velocity of the teams. A platform should support the cross-functional team in practicing DevOps: building the product features and delivering them to customers quickly and with minimum risk of causing defects or outages.

@@

It is important to keep in mind that the overall goal of using a product platform is to reduce the work not directly related to delivering customer value to a minimum. It is easy for a product platform to turn from an enabler into a source of impediments or even blockers and great care must be taken to address this. As examples, consider platform downtime, working around (missing) platform features, repeatedly performing manual tasks or coordinating with other teams to perform tasks (we will discuss the coordination costs in the next section in more detail).

@@

Sharing the platform implicitly requires multi-tenancy features, e.g. the concept of organisations and projects, granular role-based access control, resource limits and isolation, etc. Moreover, in order for the platform to be shared by all products, the feature set needs to be a union of all the features required by each product. In the theoretical extreme cases, this means supporting $n$ disjoint sets of features for $n$ products, or forcing all the products to use the single set of features prescribed by the platform.

@@

After the initial build is completed, more investment needs to be made on an ongoing basis to run operations, i.e. provision extra computing resources, patch operating systems with fixes, clean up older instances of services, etc. In order to serve the business for several years, new features need to be added as they become available in the industry in order to stay within the recognised best practice (often referred to as evergreening).

By far the largest ongoing cost to the business, which is often not explicitly recognised, is the cost of coordination between the product teams and the platform engineering team. Unless every common task from provisioning and getting access to regular deployments and elastic scaling is self-service, the product teams need to ask for help from the platform team. The platform team quickly forms a large backlog of day to day tasks to perform, product teams end up queueing behind others, and the platform team have much less time to invest into improvements of the platform itself and automation of the daily tasks. As a result, the platform quickly falls behind compared to capabilities generally available in the industry.

@@

The overall cost of building and operating a shared product platform is large enough to pose a serious risk to the business. Typically, when things do not go well, this leads to resistance to cut the losses - the sunk cost fallacy~\cite{TheSunkC23:online}. Instead of trying again, or exploring different solutions, the attitude turns into \"we will make it work\" and results in products being built on a mediocre platform with reliability and scaling issues, increasing risk of preventable outages and slowing down product delivery. This is the final, often unrecognised, cost of a shared product platform - the missed opportunity cost of not delivering as much customer value as possible.

@@

The common theme and, in our opinion, the underlying source of the shared platform costs is the fact they are \textit{shared} among a relatively large number of different product teams. (Notice that the problems listed above seem much less significant if you imagine the platform is only supporting three product teams). The shared nature is what leads to the need for the extra features that enable the sharing, it is also the source of scaling and reliability challenges and, most significantly leads to a sharp rise of coordination costs.

@@

This is not a new finding, automating certain tasks is an accepted practice in multiple industries including software engineering. An obvious example is regression testing using automated unit tests, integration tests and end-to-end tests. Testing used to be a manual task performed by humans following test scripts. Automating it significantly decreased the time between deployments allowing us to get from deploying every few months to several times a day, enabling a completely new working practice known as Continuous Deployment to Production (CDP). We will focuse on the consequences of CDP at the end of section \ref{sec:microplatforms}.

Similarly, the success of cloud services is in a huge part down to enabling automation via infrastructure provisioning APIs. Instead of spending weeks manually provisioning servers and network infrastructure, the same infrastructure can be provisioned automatically within minutes. Leveraging cloud services, a small team of engineers can set up a web or mobile application infrastructure that will vastly outperform one set up manually (over weeks or months) by a traditional IT organisation in every measure. [ref?]

@@

An important consequence of the way modern cluster management software is controlled is the ability to provide \textbf{service orchestration as code}, i.e. the state of the cluster is configured down to the desired microservices that should run on it and their names, number of replicas and deployment strategies. Coupled with service discovery, this makes it significantly easier to operate applications built with a microservices architecture.

@@

GitOps

@@

It removes the coordination cost of all product teams depending on a single platform team. If a team needs to change some aspect of their platform or add a missing feature, they can do so for themselves. A platform team will most likely still exist, but the product it delivers is the design and automation of a platform, not the infrastructure itself. This team can become an owner and steward of the platform design and automation code, which can and should be internally open source, so that everyone can contribute and the platform engineers can transition into expert support and innovation roles, rather than day to day operations. This is a tried and tested approach which works well for shared, reusable code libraries and services []. With the shared dependency removed, the organisation becomes more horizontally scalable, just like software systems do. Vertical integration leads to horizontal scalability.

@@

Targeting a cluster orchestration technology, applications become fully portable between any infrastructure supporting the technology from a developer’s laptop through any public cloud provider to on-premise solutions. In the past year, Kubernetes [] has rapidly become a de-facto standard abstraction in cluster orchestration: it is possible to run a local cluster using a tool like Minikube [], clusters can be provisioned using simple compute resources in different clouds with Kops [], all three major cloud providers are offering or have announced hosted services (GKE [], EKS [] and AKS []) and managed on-premise solutions also exist, e.g. OpenShift by RedHat []or Pivotal Kubernetes Service [].

Finally, the level of automation required for a small platform to be managed by an average sized product team enables a capability crucial to agile software delivery: continuous deployment to production.

With continuous deployment to production, every change to the application and infrastructure goes through thorough automated testing and validation. Once the validation passes without issues, the change is immediately deployed to production. This can happen several times a day, even multiple times an hour.

@@

Immutability, i.e. preventing things from being updated in place, implicitly creates a version history of all changes and makes it impossible for anything to change unexpectedly or get into a certain state in any other way than being recreated. Removing previous versions of things is a case of garbage collection to reclaim resources they consume. Immutability decouples values, i.e. the entities (infrastructure, build artifacts, deployments, etc.) from the identities, the way we refer to them over time, e.g. “the customer service”. [Rich Hickey]

@@

With immutability and with content addressing especially, it is much easier and efficient to compare two items of the same type (i.e. just by their names), enabling idempotence: the concept of operations which can be repeated many times producing no further action beyond the first application. Enabling easy idempotence leads to huge simplification, as we can typically stop focusing on detail and let idempotence handle the unwanted repetition. In a sense, idempotence unifies creation and update and makes both operations the same from outside, moving solutions from imperative to declarative. We can say that idempotence decouples intent from current state. This also appears multiple times: infrastructure provisioning with Terraform, Docker image builds and pushes to registry, deployment to a cluster or response to failure with the reconciliation loop. Everything tries to reach a declared state.
